---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
```{r}
library(ggplot2)
```

## Loading and preprocessing the data
```{r}
if (!file.exists("activity.csv")) {
  unzip("activity.zip")
}

activity <- read.csv("activity.csv", stringsAsFactors = FALSE)

activity$date <- as.Date(activity$date, format = "%Y-%m-%d")
```

## What is mean total number of steps taken per day?

Calculate the total steps taken each day.
```{r}
totalStepsByDate <- aggregate(steps ~ date, data = activity, sum, na.action = na.omit)
```

Plot the histogram.
```{r}
print(
  qplot(
    x = steps, 
    data = totalStepsByDate, 
    geom = "histogram", 
    color = I("black"), 
    fill = I("royalblue3"), 
    binwidth = 400, 
    main = "Histogram of steps taken per day.", 
    xlab = "Total Daily Steps", 
    ylab = "Frequency"
  )
)
```

The mean of the total number of steps is:
```{r}
mean(totalStepsByDate$steps)
```

The median of the total number of steps is:
```{r}
median(totalStepsByDate$steps)
```


## What is the average daily activity pattern?

Calculate the mean steps for each interval.
```{r}
averageStepsByInterval <- aggregate(steps ~ interval, data = activity, mean, na.action = na.omit)
```

Plot the series.
```{r}
print(
  qplot(
    interval, 
    steps, 
    data = averageStepsByInterval, 
    geom = "line", 
    main = "Average number of steps taken by 5-minute intervals", 
    xlab = "5-minute Interval", 
    ylab = "Average Steps"
  )
)
```

The 5-minute interval with the maximum steps on average is:
```{r}
averageStepsByInterval[averageStepsByInterval$steps == max(averageStepsByInterval$steps), ]
```

## Imputing missing values

Total number of missing values:
```{r}
sum(is.na(activity$steps))
```

**Imputation Method**

Replace missing values with the mean for the 5-minute interval. 

1. Merge original and average datasets
2. Replace NA values with mean value.

Step 1: Merge activity and averageStepsByInterval datasets
```{r}
activityImputed <- merge(activity, averageStepsByInterval, by.x = "interval", by.y = "interval")
names(activityImputed) <- c("interval", "steps", "date", "meanStepsByInterval")
```

Step2: Replace NA values with mean value.
```{r}
activityImputed$steps[is.na(activityImputed$steps)] <- activityImputed[is.na(activityImputed$steps), "meanStepsByInterval"]
```

As before, show histogram, mean and median, this time with imputed data.
```{r}
totalStepsByDateImputed <- aggregate(steps ~ date, data = activityImputed, sum)

print(
  qplot(
    x = steps, 
    data = totalStepsByDateImputed, 
    geom = "histogram", 
    color = I("black"), 
    fill = I("seagreen4"), 
    binwidth = 400, 
    main = "Histogram of steps taken per day.", 
    xlab = "Total Daily Steps", 
    ylab = "Frequency"
  )
)
```

The mean of the total number of steps is:
```{r}
mean(totalStepsByDateImputed$steps)
```
---
The median of the total number of steps is:
```{r}
median(totalStepsByDateImputed$steps)

```


## Are there differences in activity patterns between weekdays and weekends?

Add daytype factor column indicating if the date is a weekday or weekend.
```{r}
activityImputed$daytype <- factor(ifelse(weekdays(activityImputed$date) %in% c("Saturday", "Sunday"), "weekend", "weekday"))
```

Take the average steps by interval and day type.
```{r}
averageStepsByIntervalAndDayType <- aggregate(steps ~ interval + daytype, data = activityImputed, mean)
```

Plot the two day type series.
```{r}
print(
  qplot(
    interval, 
    steps, 
    data = averageStepsByIntervalAndDayType, 
    color = daytype, 
    facets = daytype ~ ., 
    geom = "line", 
    main = "Average number of steps taken by 5-minute intervals and day type", 
    xlab = "5-minute Interval", 
    ylab = "Average Steps"
  )
)
```